---
title: "Drought forecasting exercise"
author: "Jordan Richards"
date: "`r Sys.Date()`"
output: beamer_presentation
---
  
# Dataset
  
* Exercise developed by Chris Wikle and Dan Pagendam (2019).
* The data consists of:
  * monthly 33 x 84 grids (2 degree x 2 degree) of sea surface temperature (SST) anomaly (2772 pixels).
  * monthly rainfall anomaly in mm for the Murray Darling Basin (MDB).
* Obtained from two sources:
  * http://www.bom.gov.au/climate/change/
  * http://iridl.ldeo.columbia.edu/
* We will use a type of recurrent NN(LSTM) model to obtain 3-month-out forecasts of rainfall anomaly using SST grids as a predictor.

# Required packages

We will be using some functions and the images from Dan's github directory, https://github.com/dpagendam/deepLearningRshort

```{r,eval=T}
#remotes::install_github("dpagendam/deepLearningRshort")

library(keras)
library(raster)
library(deepLearningRshort)
data(drought)
```


# Visualise SST

```{r, eval=TRUE,echo=F}
plot(anomalyRasterList[[1]], main = paste0("Sea Surface Temperature Anomaly (", dateGrid[1, 1], "/", dateGrid[1, 2], ")"),
  xlab = "Longitude", ylab = "Latitude", zlim = c(-5, 5))
  
```


# Strategy

* We could apply a CNN layer recurrently to extract both sequential and spatial information from the SST grids, but this will require lots of processing power and parameters
* Instead, we treat the SST as a multivariate time series and use regular RNNs. However, we have 2772 locations, so we first reduce the dimensionality using EOFs (PCA)

# Data manipulation

```{r, eval=TRUE}
batchSize <- 32
forecastMonthsAhead <- 3
timestepsPerSample <- 24
trainingInds <- 1:1300
validationInds <- 1301:1434
# We consider only 1434 months
```

* We will project the 2772 pixels onto 100 EOFs.

```{r, eval=TRUE}
numComponents <- 100
EOFList <- rasterToEOFs(anomalyRasterList[trainingInds],
numComponents = numComponents, plot = FALSE)
v.train <- EOFList[["rasterEOFs"]][["v.dim.red"]]
```

# Plot EOFs

```{r,eval=T,echo=F}
EOFList <- rasterToEOFs(anomalyRasterList[trainingInds], numComponents = numComponents, plot = TRUE)

```

# Plot EOFs

```{r,eval=T,echo=F}
validPixels <- EOFList[["raster.validPixels"]]
eof1 <- EOFsToRaster(matrix(EOFList$rasterEOFs$EOFs[, 1], ncol = 1),
        matrix(rep(1, 1), nrow = 1), c(33, 84),
        validPixels)[[1]]
extent(eof1) <- extent(anomalyRasterList[[1]])
plot(eof1, main = "1st EOF", xlab = "Longitude", ylab = "Latitude")
```

# Plot EOFs

```{r,eval=T,echo=F}
eof2 <- EOFsToRaster(matrix(EOFList$rasterEOFs$EOFs[, 2], ncol = 1),
        matrix(rep(1, 1), nrow = 1), c(33, 84),
        validPixels)[[1]]
extent(eof2) <- extent(anomalyRasterList[[1]])
plot(eof2, main = "2nd EOF", xlab = "Longitude", ylab = "Latitude")
```
# Plot EOFs

```{r,eval=T,echo=F}
eof100 <- EOFsToRaster(matrix(EOFList$rasterEOFs$EOFs[, 100], ncol = 1),
        matrix(rep(1, 1), nrow = 1), c(33, 84),
        validPixels)[[1]]
extent(eof100) <- extent(anomalyRasterList[[1]])
plot(eof100, main = "100th EOF", xlab = "Longitude", ylab = "Latitude")
```

# Reconstructing SST


```{r, eval=TRUE}
validationSample <- 1434
X <- EOFList$rasterEOFs$EOFs
r1 <- anomalyRasterList[[validationSample]]
validPixels <- EOFList[["raster.validPixels"]]
Y <- getValues(r1)
Y <- Y[validPixels]
lm1 <- lm(Y~X)
intercept <- coefficients(lm1)[1]
alpha <- coefficients(lm1)[1]
beta <- coefficients(lm1)[2:(numComponents + 1)]
r2 <- alpha + EOFsToRaster(X, matrix(beta, nrow = 1),
      c(33, 84), validPixels)[[1]]
extent(r2) <- extent(r1)
```

# Reconstructing SST

```{r,eval = T, echo = F}
par(mfrow = c(1, 2))
plot(r1, xlab = "Longitude", ylab = "Latitude", zlim = c(-5, 5), main="Original")
plot(r2, xlab = "Longitude", ylab = "Latitude", zlim = c(-5, 5), "Reconstructed")
```

# Dimension reduction for validation data

* Here we project the validation data SST anomaly grids onto the same EOFs generated from the training data.
*·* You can think of v.validation as a multivariate time series of coefficients that we can use to reconstruct SST anomaly from the EOFs.

```{r,eval = T, echo = T}
v.validation <- proj.raster.EOFs(
  anomalyRasterList[validationInds],
         EOFList[["rasterEOFs"]][["EOFs"]],
         EOFList[["raster.validPixels"]])
```

# Dimension reduction for validation data


```{r,eval = T, echo = F}
par(mfrow = c(3, 1), mar = c(4,4,1,1))
plot(trainingInds, v.train[, 1], ty = "l", xlim = c(0, 1434),
     xlab = "Months", ylab = "Variable 1")
lines(validationInds, v.validation[, 1], col = "blue")
legend("topleft", legend = c("train", "validation"), horiz = TRUE,
       box.lwd = 0, col = c("black", "blue"), lty = 1)
plot(trainingInds, v.train[, 2], ty = "l", xlim = c(0, 1434),
     xlab = "Months", ylab = "Variable 2")
lines(validationInds, v.validation[, 2], col = "blue")
legend("topleft", legend = c("train", "validation"), horiz = TRUE,
       box.lwd = 0, col = c("black", "blue"), lty = 1)
plot(trainingInds, v.train[, 3], ty = "l", xlim = c(0, 1434),
     xlab = "Months", ylab = "Variable 3")
lines(validationInds, v.validation[, 3], col = "blue")
legend("topleft", legend = c("train", "validation"), horiz = TRUE,
       box.lwd = 0, col = c("black", "blue"), lty = 1)
```

# Data wrangling

* All predictors combined together

```{r,eval = T, echo = T}
v.combined <- rbind(v.train, v.validation)
```

* and normalised 

```{r,eval = T, echo = T}
v.scaling.train <- scaleCols.pos(
  v.combined[trainingInds, ])
v.train.scaled <- v.scaling.train[["X.scaled"]]
v.scaling.validation <- scaleCols.pos(
  v.combined[validationInds, ],
              colMaxsX = v.scaling.train[["colMaxsX"]],
              colMinsX = v.scaling.train[["colMinsX"]])
v.validation.scaled <- v.scaling.validation[["X.scaled"]]

v.scaled <- rbind(v.train.scaled, v.validation.scaled)
```

# Formatting data for an RNN

```{r, eval = T}
numDims <- ncol(v.scaled)
tensorData <- tensorfyData.rnn(v.scaled, 
              forecastMonthsAhead,
              timestepsPerSample, indicesX = 1:numDims,
              indicesY = 1:numComponents, 
              indicesTrain = trainingInds,
              indicesTest = validationInds)
str(tensorData)
```

* Currently Y.train is a sample of sequences of MV SST anomalies, but we want to forecast the 3-month out rainfall anomaly

# Response data

```{r}
Y.train.inds <- tensorData$y.train.tsInds
Y.valid.inds <- tensorData$y.test.tsInds
Y.train.rnn_MDB <- rainfallAnomaly[Y.train.inds, 3]
Y.valid.rnn_MDB <- rainfallAnomaly[Y.valid.inds, 3]
```

* We will also normalise the response, but this is only because we will fit a Gaussian model

```{r}
Y.train.min <- min(Y.train.rnn_MDB)
Y.train.max <- max(Y.train.rnn_MDB)

Y.rnn.train <- (Y.train.rnn_MDB - Y.train.min)/
                    (Y.train.max - Y.train.min)
Y.rnn.valid <- (Y.valid.rnn_MDB - Y.train.min)/
                    (Y.train.max - Y.train.min)
```              
                    
# RNN tensors                

We finally get

```{r}
X.rnn.train <- tensorData[["X.train.rnn"]]
X.rnn.valid <- tensorData[["X.test.rnn"]]

dim(X.rnn.train)
length(Y.rnn.train)

```

# Custom loss

```{r}
Gaussian_logLikelihood <- function(y_true, y_pred)
{
  K <- backend()
  # Extract the first and second columns of predictions
  mu <- (y_pred[,1])
  sigma <- K$exp(y_pred[,2])
  
#Extract first column of y_true to ensure same dimension
  y <- y_true[,1]
  
  ll <- -0.5*((mu - y)/(sigma))^2 - K$log(sigma)
  ll <- ll -0.5*K$log(2*pi)
  
  return( -(K$sum(ll)))
}
```


# Building an LSTM Model 

```{r, eval = T}
library(keras)
input.lay <- layer_input(shape = dim(X.rnn.train)[-1], 
      name = 'input_layer')

model <- input.lay %>%
  layer_lstm(units = 128,
      input_shape = c(timestepsPerSample, numDims),
      return_sequences = FALSE) %>%
  layer_dense(units = 128, activation = "relu") 
  output.lay <- model %>% layer_dense(units = 2)
```

# Compile

```{r, eval=TRUE}


  model <- keras_model(
    inputs = c(input.lay), 
    outputs = c(output.lay)
  )
model %>% compile(loss = Gaussian_logLikelihood,
                  optimizer = optimizer_rmsprop())

```

# Summary 

```{r, eval = F}
summary(model)
```

# Model training

Let's train the model with early stopping and a checkpoint


```{r, eval = T}
history <- model %>% fit(
  x = X.rnn.train, y = Y.rnn.train, 
  batch_size = batchSize, epochs = 200, shuffle = FALSE, 
  validation_data = list(X.rnn.valid, Y.rnn.valid),
  callbacks = list(
  callback_early_stopping(monitor = "val_loss", 
                          min_delta = 0, patience = 20),
  callback_model_checkpoint(filepath = "model_weights",
      verbose=0, monitor="val_loss",
      save_best_only = TRUE, save_weights_only = TRUE)))
#Then load the saved weights
model <- load_model_weights_tf(model,
                               filepath="model_weights")
```

# Checking performance

* We can calculate the mean and standard deviations of the 3 month out (Gaussian)
predictive distributions.
* Then create 50% and 95% prediction intervals.

```{r}
lstmPredictions <- model %>% predict(X.rnn.valid)

mu <- Y.train.min + 
  lstmPredictions[, 1]*(Y.train.max - Y.train.min)
sigma <- exp(lstmPredictions[, 2])*
  (Y.train.max - Y.train.min)
n <- length(mu)
upper95 <- mu + 1.96*sigma
lower95 <- mu - 1.96*sigma
upper50 <- mu + 0.674*sigma
lower50 <- mu - 0.674*sigma
```


# Checking performance

Plot the true time series with 3-month-out forecast and 50% and 95%
prediction intervals.

```{r, eval = F}
plot(rainfallAnomaly[Y.validation.inds, 3], ty = "l", 
     xlab = "Time (months)", 
     ylab = "MDB Rainfall Anomaly (mm)")
lines(mu, col = "blue")
polygon(x = c(1:n, rev(1:n), 1),
        y = c(lower95, rev(upper95), lower95[1]),
        col = fade("blue", 100), border = NA)
polygon(x = c(1:n, rev(1:n), 1),
        y = c(lower50, rev(upper50), lower50[1]),
        col = fade("blue", 100), border = NA)
```

# Checking performance


```{r, eval = T, echo = F}
plot(rainfallAnomaly[Y.valid.inds, 3], ty = "l", 
     xlab = "Time (months)", 
     ylab = "MDB Rainfall Anomaly (mm)")
lines(mu, col = "blue")
polygon(x = c(1:n, rev(1:n), 1),
        y = c(lower95, rev(upper95), lower95[1]),
        col = fade("blue", 100), border = NA)
polygon(x = c(1:n, rev(1:n), 1),
        y = c(lower50, rev(upper50), lower50[1]),
        col = fade("blue", 100), border = NA)
```

# Checking performance

Calculate what percentage of the time the true rainfall anomaly was within
the 50% and 95% prediction intervals.

```{r, eval = T}
n <- (length(Y.valid.inds))
coverage50 <- length(
  which(rainfallAnomaly[Y.valid.inds, 3]> lower50
  & rainfallAnomaly[Y.valid.inds, 3] < upper50))/n
coverage95 <- length(
  which(rainfallAnomaly[Y.valid.inds, 3] > lower95
  & rainfallAnomaly[Y.valid.inds, 3] < upper95))/n
print(coverage50)
print(coverage95)

```


# Extensions

* How does varying the number of units in the LSTM layer affect the predictions?
* How do the predictions change if you add three dense layers after the LSTM layer (instead of just 1)?
* How are the predictions if much fewer EOFs are used for prediction?
* How does fewer EOFs affect the number of parameters in the model?
  
  
  